\chapter{Background and problem}

% - old ud-gf
% - newer naive approach - github + article 2017
%   - limitations
%   - performance



% \section{Why these three problems?}
% \todo[inline]{Because they were issues that popped up during the work at SMU}
% 
% I have things to write here but I don't know what those things are. 

% Background to the assignment. Why is it relevant?
\section{Background}\label{sect:background}
% Application in 

In a project\cite{listenmaa-etal-2021-towards} at the Singapore Management University, Center for Computational Law we wanted to convert text into logical form, with the goal of being able to parse and interpret legal text. We initially tried using GF directly, but we ran into issues with it being too strict, so many of the sentences were not parsed at all, either because they contained words not in the dictionary or because it didn't follow the grammatical rules exactly. In order to work around this we used a UD parser, which was based on machine learning, which allowed it to be more robust for unexpected data. However, now the problem of converting into logic became more difficult, so we used the gf2ud tool to convert the UD trees into GF trees. This worked fairly well, but we ran into several problems, described in the next section, which this project aims to solve.

\section{Problem}
% The formulation of the problem at hand and, the assignment. This should include an extended version of the scientific problem definition and references to knowledge within the area given in the thesis proposal.

% What has been done before and what remains to be done

% 2. Background & Problem
% - old ud-gf
% - newer naive approach - github + article 2017
%   - limitations
%   - performance

% \todo[inline]{Rewrite this to not be for planning report}

The original implementation of ud2gf had some limitations. There are primarily three problems that this work aims to fix:

\begin{enumerate}
    \item Converting UD trees to GF trees quickly became extremely slow for sentences with more than a couple of words and/or
    when using large GF-grammars, e.g. the GF-grammars based on Wordnet\cite{angelov2016predicting}.
\item
    If the structure would differ too much between the representation of a sentence in UD format and as a GF tree, it is not possible to describe the required transformation in the old ``labels file'' language. See \autoref{improving-flexibility-macros} for more details.
\item
    The third problem is that it can sometimes be difficult to figure out why a rule in a labels-file is not firing, so it would be useful to have a debugging tool to help diagnosing such issues.
\end{enumerate}


% Ideas:
%
% - Limits on how to transform trees - solution: extend language. currently hack using lambda calculus, but maybe better macros
%
% - Evaluate effectiveness of tool
%
% - (somewhere mention the performance boosts and analyse the complexity of it)

% This section is optional. It may be used if there is a need to describe the problem that you want to solve in more technical
% detail and if this problem description is too extensive to fit in the introduction.

% From elsewhere:
% - Background: GF \cite{ranta-2004}, UD \cite{nivre-etal-2016-universal},
%   - previous work: gf2ud \cite{kolachina-ranta-2016}, ud2gf \cite{kolachina-ranta-2017}
% - Describe the new algorithm
% - Extending the macro language
%   - Need for improvement: to match "fluffy and cute" needs 2 levels of nesting
%   - Solution: continuations
% - Case study: legal language


% 3. The new algorithm
% - definitions
% - examples
% - how annotations work

% \section{Context}
% \todo[inline]{What's the difference between this and intro?}
% This work is mostly a continuation of the work in \cite{kolachina-ranta-2017}.

% A practical problem for which this tool can be useful can be found in \cite{listenmaa-etal-2021-towards}. In the Future Work section, under "Robust fall-back options", gf2ud is mentioned as a possible solution to making the parser more robust.

% % Use one or two relevant and high quality references for providing evidence from the literature that the proposed study indeed
% % includes scientific and engineering challenges, or is related to existing ones. Convince the reader that the problem addressed
% % in this thesis has not been solved prior to this project.

% Aim for the work. What should be accomplished?
\section{Goals and Challenges}

% try to preserve as much information as possible from the UD tree

% do this in an efficient way

% \todo[inline]{No longer relevant for final report?}

% \todo[inline]{Change this so it's clear that it has already been done}

\begin{enumerate}
    \item 
Analyze algorithms and improve performance. The main challenge here is to find an algorithm for finding matching trees without exponential complexity on the number of children of a node in the UD tree.

\item
% macro -> annotation
Improve flexibility of annotation language, allowing changing the structure of the trees while translating from UD to GF. One challenge here is in figuring out either how to change the algorithms to support these more advanced transformations or to find a way to allow them without needing to change the algorithms.
\item
 Write a debugging tool, which analyzes exactly what it is that prevents a rule in a labels-file from firing or what prevents that tree from being selected.
One challenge here is how to explain to the user the issue for all the possible things that can go wrong.
There is also an engineering challenge in making an algorithm that figures out what went wrong and why.
% either repeating the steps of the algorithm in order to find the issues or transforming the algorithm in a way that they can explain what went wrong

\item
 Document the version of the tool on which this work is based on, which had changed since what was written in \cite{kolachina-ranta-2017}
\end{enumerate}

 % improve

% % Analyse the complexity of old and new algorithm, evaluate effectiveness of ...

% % Describe your contribution with respect to concepts, theory and technical goals. Ensure that the scientific and engineering
% % challenges stand out so that the reader can easily recognize that you are planning to solve an advanced problem.


% Limitations. What should be left out and why?
\section{Limitations}

Only the direction of converting UD trees to GF trees is studied here, because that is what was relevant to the application at hand. Furthermore, the two directions are almost completely independent in the implementation.

% I normalfallet behöver avgränsningarna inte motiveras.

% Limit the work mostly to what has already been done

% Har gjort:
%

%  \todo{why?}

% \todo[inline]{other limitations?}
